---
output:
  pdf_document: default
  html_document: default
  word_document: default

---

\def\by{\mbox{\boldmath$y$}}
\def \lam{\lambda}
\def\bt{\mbox{\boldmath$t$}}
\def \ds{\displaystyle}
\newcommand{\hg}[2]{\mbox{}_{\scriptscriptstyle #1} F_{\scriptscriptstyle #2}}

# FUNCTIONS

The R code in this section of the Web Appendix is for functions which generate data and which evaluate the log-likelihoods for the $N$-mixture models introduced in the main paper. The  abundance is taken to be site-dependent or constant and the rate parameter and search time to be constant. The functions are called from the R code for the simulations presented in the program section. Each set of functions for a given model is prefaced by the formula for the log-likelihood of that model. 

## Model M1s: Detection/non-detection with a single visit


### Formula 


For $R_1$ the number of sites at which a detection is recorded,  
$$
L_{M1s} (h, \lam_i;\by)=  \prod_{i=1}^{R_1} \big\{ 1- \exp[-\lam_i (1 - e^{-h T})]  \big\}\prod_{i=R_1+1}^{R} \exp[-\lam_i (1 - e^{-h T})]
$$
where the data $\by=(y_1, \ldots, y_R)$ is a binary vector.
If $\lam_i$ is a constant, the unknown parameters $\lam$ and $h$ are not identifiable.


### R code

``` {r eval=FALSE}
# generates  detection/non-detection data for M1s with lambda site-dependent
xgM1sgen=function(param)
{
  b0=param[1]
  b1=param[2]
  h=param[3]
  lambda=exp(b0+b1*vegHt)
  yvec=matrix(0,R,1)
  n = rpois(R, lambda)
  pvec=1-exp(-h*n*Tmax)
  for(i in 1:R)
   {yvec[i]=rbinom(1,1,pvec[i])}
  data=cbind(yvec)
  return(data)
}

# evaluates the log-likelihood for M1s with lambda site-dependent
xgM1s=function(param)
{
  b0=param[1]
  b1=param[2]
  h=exp(param[3])
  lam=as.matrix(exp(b0+b1*vegHt))
  yvec=gdatx
  loglik=sum(-(1-yvec)*lam*(1-exp(-h*Tmax))+
                yvec*log(1-exp(-lam*(1-exp(-h*Tmax)))) )
  return(-loglik)
}
 
```

## Model M1m: Detection/non-detection with multiple visits


### Formula 


For $R_1$ the number of sites and $J_{i,1}$ the number of visits within site $i$ at which  detections are recorded,
\begin{eqnarray*}
L_{M1m}(h, \lam_i;\by) & = &  \prod_{i=1}^{R_1} \Big[ \sum_{j=0}^{J_{i,1}} \binom{J_{i,1}}{j} (-1)^j \exp(\lam_i  e^{\,- j  h T}) \Big] \times \prod_{i=1}^R  \exp[-\lam_i (1- e^{-  \, h T J_{i,0}})] 
\end{eqnarray*}
where $J_{i,0}=J - J_{i,1}$, $i=1, \ldots, R_1$,  $J_{i,0}=J, i=R_1+1, \ldots, R$ and $\by$ is the vector of binary responses, arranged conformably. 
The formula for constant $\lam$ follows immediately.


### R code
``` {r eval=FALSE}
# generates detection/non-detection data for M1m with lambda constant
gM1mgen=function(param)
{
  lam=param[1]
  h=param[2]
  ymat=matrix(0,R,J)
  for(i in 1:R)
   {n = rpois(1, lam)
    p=1-exp(-h*n*Tmax)
    ymat[i,]=rbinom(J,1,p)
  }
  return(ymat)
}

# evaluates the log-likelihood for M1m with lambda constant
gM1m=function(param)
{
  lam=exp(param[1])
  h=exp(param[2])
  J1=rowSums(gdat) # number of detections at the R sites
  J0=J-J1
  loglik=matrix(0,R,1)
  for(i in 1:R)
  {
  J1i=J1[i];
  J0i=J0[i];
  term=0;
  for(j in 0:J1i)
  {a=exp(-h*Tmax*(j+J0i));
  term=term+(-1)^j*choose(J1i,j)*(exp(lam*a))
  }
  loglik[i]=-lam+log(term)+log(choose(J,J1i))
  }
  return(-sum(loglik))
}

# generates detection/non-detection data for M1m with lambda site-dependent
xgM1mdat=function(param)
{
  b0=param[1]
  b1=param[2]
  h=param[3]
  lambda=exp(b0+b1*vegHt)
  ymat=matrix(0,R,J)
  for(i in 1:R)
  {n = rpois(1,lambda[i])
  p=1-exp(-h*n*Tmax)
  for(j in 1:J)
  {ymat[i,j]=rbinom(1,1,p)}
  }
  return(ymat)
}

# evaluates the log-likelihood for M1m with lambda site-dependent
xgM1mT=function(param)
{b0=param[1]
b1=param[2]
h=exp(param[3])
J1=rowSums(gdatx)
J0=J-J1
lam=exp(b0+b1*vegHt)
loglik=matrix(0,R,1)
# note that if J1[i]=0 the expresson reduces correctly to
# exp(-lam[1-e^(-hTmax J)])
for(i in 1:R)
{ J1i=J1[i];
J0i=J0[i];
term=0;
for(j in 0:J1i)
{a=exp(-h*Tmax*(j+J0i));
term=term+(-1)^j*choose(J1i,j)*(exp(lam[i]*a))
}
loglik[i]=-lam[i]+log(term)+log(choose(J,J1i))
}
return(-sum(loglik))
}
```

## Model M2s: Time-to-first-detection with a single visit


### Formula 

For $R_t$ the number of sites at which a time-to-first-detection is recorded,
$$
L_{M2s} (h, \lam_i; \bt_{(1)}) = \prod_{i=1}^{R_t}  \lam_i \, h \exp [ -\lam_i (1 - e^{-h t_{i(1)}}) - h t_{i(1)} ] \times \prod_{i=R_t+1}^{R} \exp[-\lam_i ( 1- e^{-h T} )] 
$$
where $\bt_{(1)}$ is the vector of times. The formula for constant $\lam$ follows immediately with the parameters $h$ and $\lam$ identifiable.

### Rcode

```{r eval=FALSE}
# generates times-to-first-detection for M2s with lambda constant
# returns a vector with no detection indicated by Tmax
gM2sgen=function(param)
{lam=param[1]
h=param[2]
tvec=matrix(0,R,1)
for(i in 1:R)
{n=rpois(1,lam);
if(n==0)
{tvec[i]=Tmax}
if(n > 0)
{ti=rexp(1,h*n);
if(ti>Tmax)
{tvec[i]=Tmax}
if(ti<Tmax)
{tvec[i]=ti}
}
}
return(tvec)
}

# evaluates the log-likelihood for M2s with lambda constant
gM2s=function(param)
{lam=exp(param[1])
h=exp(param[2])
ti=as.matrix(subset(gdat,gdat < Tmax))
Rt=nrow(ti) # total number of sites at which t_i1 < Tmax
RT=R-Rt
sumt=sum(ti)
loglik=-RT*lam*(1-exp(-h*Tmax))+
     Rt*log(h*lam) -lam*sum(1-exp(-h*ti))-h*sumt
return(-loglik)
}

# generates times-to-first-detection for M2s with lambda site-dependent
xgM2sgen=function(param)
{b0=param[1]
b1=param[2]
h=param[3]
lam=exp(b0+b1*vegHt)
tvec=matrix(0,R,1)
for(i in 1:R)
  {n=rpois(1,lam[i]);
   if(n==0)
    {tvec[i]=Tmax}
   if(n > 0)
    {ti=rexp(1,h*n);
      if(ti>Tmax)
       {tvec[i]=Tmax}
      if(ti<Tmax)
      {tvec[i]=ti}
}
}
return(tvec)
}

# evaluates the log-likelihood for M2s with lambda site-dependent
xgM2s=function(param)
{b0=param[1]
b1=param[2]
h=exp(param[3])
lam=exp(b0+b1*vegHt)
loglikt=0
loglikT=0
for(i in 1:R)
{
  if(gdatx[i] < Tmax)
  {loglikt=loglikt+log(h*lam[i])-lam[i]+ lam[i]*exp(-h*gdatx[i])-
       h*gdatx[i]}
  if(gdatx[i] == Tmax)
  {loglikT=loglikT-lam[i]*(1-exp(-h*Tmax))}
}
loglik=sum(loglikt+loglikT)
return(-loglik)
}
```

## Model M2m: Time-to-first-detection with multiple visits

### Formula 

For $R_t$ the number of sites and  $J_{i,t}$ the number of visits within site $i$ at which  times-to-first-detection are recorded,  
\begin{eqnarray*}
L_{M2m} (h, \lam_i; \bt_{(1)}) & = &  \ds \prod_{i=1}^{R_t} \bigg\{  h^{J_{i,t}}   \exp \left [-\lam_i (1-e^{-W_{i,t}} ) \right ]  \ds \sum_{k=0}^{J_{i,t}} (\lam_i e^{-W_{i,t}})^k  S(J_{i,t},k) \bigg\} \\
& & \hspace{15mm} \times  \ds \prod_{i=R_t+1}^{R}  \exp[-\lam_i (1-e^{- h T J})] 
\end{eqnarray*} 
where $\bt_{(1)}$ is the vector of times, arranged conformably,  $W_{i,t} = \sum_{j=1}^{J_{i,t}} h t_{i(1)}+ (J-J_{i,t}) h T, i=1, \ldots, R_t$, and $S(J_{i,t},k)$ is a Stirling number of the second kind. The formula for constant $\lam$ follows immediately.


### R code

``` {r eval=FALSE}
# generates times-to-first-detection for M2m with lambda constant
gM2mgen=function(param)
{lam=param[1]
h=param[2]
tmat=matrix(0,R,J)
for(i in 1:R)
{n=rpois(1,lam)
if(n==0) {tmat[i,]=Tmax}
if(n >0)
{tvec=rexp(J,h*n)
tvec[tvec>Tmax]<-Tmax
tmat[i,]=tvec
}
}
return(tmat)
}

# evaluates the log-likelihood for M2m with lambda constant
gM2m=function(param)
{lam=exp(param[1])
h=exp(param[2])
Jt=matrix(0,R,1)
W=matrix(0,R,1)
for(i in 1:R)
{Jt[i]=length(subset(gdat[i,], gdat[i,] < Tmax))
# note that W[i] is a sum of all times at site i
# including those recorded as Tmax, that is ti1 >Tmax
W[i]=sum(gdat[i,])}
mu=lam*exp(-h*W)
fin=matrix(0,R,1)
for(i in 1:R)
{
  if(Jt[i] > 0)
  {
    term=0
    for(k in 1:Jt[i])
    {
    term=term+(mu[i]^k)*Stirling2(Jt[i],k)}
    fin[i]=fin[i]+log(term)
  }
}
loglik=-R*lam+sum(mu)+sum(Jt)*log(h)+
  sum(log(choose(J,Jt)))+sum(fin)
return(-loglik)
}

# generates times-to-first-detection for M2m with lambda site-dependent
xgM2mgen=function(param)
{b0=param[1]
b1=param[2]
h=param[3]
lam=exp(b0+b1*vegHt)
tmat=matrix(0,R,J)
for(i in 1:R)
{N=rpois(1,lam[i])
if(N==0) {tmat[i,]=Tmax}
if(N >0)
{tvec=rexp(J,h*N)
tvec[tvec>Tmax]<-Tmax
tmat[i,]=tvec
}
}
return(tmat)
}

# evaluates the log-likelihood for M2m with lambda site-dependent
xgM2m=function(param)
{b0=param[1]
b1=param[2]
h=exp(param[3])
lam=exp(b0+b1*vegHt)
Jt=matrix(0,R,1)
W=matrix(0,R,1)
for(i in 1:R)
{Jt[i]=length(subset(gdatx[i,], gdatx[i,] < Tmax))
# note that W[i] is a sum of all times at site i
# including those recorded as Tmax, that is ti1 >Tmax
W[i]=sum(gdatx[i,])}
mu=lam*exp(-h*W)
fin=matrix(0,R,1)
for(i in 1:R)
{
  if(Jt[i] > 0)
  {
    term=0
    for(k in 1:Jt[i])
    {
    term=term+(mu[i]^k)*Stirling2(Jt[i],k)}
    fin[i]=fin[i]+log(term)
  }
}
# note that the term -lam_i (1-exp(-h W_i)) = -lam_i+[lam_i exp(-h W_i)]
# covers all R since if Jt[i]=0 then W[i] = h*Tmax*J
loglik=-sum(lam)+sum(mu)+sum(Jt)*log(h)+
  sum(log(choose(J,Jt)))+sum(fin)
return(-loglik)
}
```
 
## Model M3s: Counts with a single visit

### Formula 

The distribution of counts at a given site follows a thinned Poisson. For $R$ sites,
$$
L_{M3s}(h,\lam_i; \by) = \prod_{i=1}^R \frac{[\lam_i (1- e^{-h T })]^{y_i} \exp
[-\lam_i (1- e^{-h  T}) ]}{y_i!},
$$
where $\by =(y_1, \ldots, y_R)$ is the vector of counts. If $\lam_i$ is a constant, the unknown parameters $\lam$ and $h$ are not identifiable.


### R code

``` {r eval=FALSE}
# generates binomial count data for M3s with lambda site-dependant
xgM3Bsgen=function(param)
{b0=param[1]
b1=param[2]
h=param[3]
lam=exp(b0+b1*vegHt)
p=1-exp(-h*Tmax)
yvec=matrix(0,R,1)
for(i in 1:R){
  n=rpois(1,lam[i])
  yvec[i]=rbinom(1,n,p)}
return(yvec)
}

# evaluates the log-likelihood for M3s with lambda site-dependant
xgM3Bs=function(param)
{b0=param[1]
b1=param[2]
h=exp(param[3])
yvec=gdatx
lam=exp(b0+b1*vegHt)
p=1-exp(-h*Tmax)
loglik=sum(yvec*log(lam))+sum(yvec*log(p))-
  sum(lam*p)-sum(log(factorial(yvec)))
return(-loglik)
}
```

## Model M3m: Counts with multiple visits 

### Formula 

For $R_1$ sites with the maximum count greater than zero, that is $y_{i,M} > 0$,
\begin{eqnarray*}
L_{M3m}(h,\lam_i;\by) & = &   \prod_{i=1}^{R_1} \bigg\{ \; e^{-\lam_i}  \; \left  (\frac{1-e^{-h T}}{e^{-h T}} \right )^{\sum_{i=1}^J y_{ij}} \; \left ( \lam_i  e^{- h T J} \right )^{y_{i,M}} \\
& & \hspace{10mm} \times    \;  \hg{J-1}{J-1} 
\bigg( 
 \begin{array}{c}
(y_{i,M}+1)\\ 
(y_{i,M}-y_{ij}+1)
\end{array}
\bigg| \lam_i  e^{-h T J}) \bigg) \bigg\}                    \\
& & \hspace{20mm} \times \prod_{i=R_1+1}^R \exp[-\lam_i (1 - \exp^{-h T J})]
\end{eqnarray*}
where $\by$ is the vector of counts, arranged conformably. The formula for constant $\lam$ follows immediately.

### R code

``` {r eval=FALSE}
# generates binomial count data for model M3m with lambda constant
gM3Bmgen=function(param)
{lam=param[1]
h=param[2]
ymat=matrix(0,R,J)
for(i in 1:R){
  n=rpois(1,lam)
  p=1-exp(-h*Tmax)
  for(j in 1:J)
  {
  ymat[i,j]=rbinom(1,n,p)}
}
return(ymat)
}

# evaluates the log-likelihood for model M3m with lambda constant
gM3Bm=function(param)
{
  lam=exp(param[1])
  h=exp(param[2])
  loglik=0
  p=1-exp(-h*Tmax)
  ymat=gdat
  for(i in 1:R)
  {
    # set up yvec, p, lograt
    yvec=ymat[i,]
    lograt=sum(yvec)*log(p/(1-p))
    #sort yvec
    yvec=sort(yvec)
    yij=yvec[-J]
    yiJ=yvec[J]
    term=sum(log(choose(yiJ,yij)))
    # calculate hypergeometric function
    theta=lam*(1-p)^J
    hgfun=genhypergeo(rep(yiJ,(J-1)) + 1,yiJ - yij + 1,theta)
    loglik=loglik-lam+lograt+yiJ*log(theta)+
      log(hgfun)+term-log(factorial(yiJ))
  }
  return(-loglik)
}


# generates binomial count data for model M3m with lambda site-dependent
xM3Bmgen=function(param)
{b0=param[1]
b1=param[2]
lam=exp(b0+b1*vegHt)
h=param[3]
ymat=matrix(0,R,J)
for(i in 1:R){
  n=rpois(1,lam[i])
  p=1-exp(-h*Tmax)
  for(j in 1:J)
  {
  ymat[i,j]=rbinom(1,n,p)}
}
return(ymat)
}

# evaluates the log-likelihood for model M3m with lambda site-dependent
xgM3Bm=function(param)
{b0=param[1]
b1=param[2]
h=exp(param[3])
lam=exp(b0+b1*vegHt)
loglik=0
ymat=gdatx
for(i in 1:R)
{
  # set up yvec, p, lograt, theta
  yvec=ymat[i,]
  p=1-exp(-h*Tmax)
  lograt=sum(yvec*log(p/(1-p)))
  theta=lam[i]*(1-p)^J
  # now sort yvec
  yvec=sort(yvec)
  yij=yvec[-J]
  yiJ=yvec[J]
  term=sum(log(choose(yiJ,yij)))
  hgfun=genhypergeo(rep(yiJ,(J-1)) + 1,yiJ - yij + 1,theta)
  # now calculate log-likelihood
  loglik=loglik+lograt-lam[i]+yiJ*log(theta)-
    log(factorial(yiJ))+term+log(hgfun)
}
return(-loglik)
}
```

## Model M4s: Times-to-each-detection with a single visit 

### Formula 

For $R_t$ the number of sites at which times-to-each-detection are recorded, the likelihood for the model simplifies to yield
$$
L_{M4s}(h,\lam_i; \by, \bt) = \prod_{i=1}^{R_t}  h^{y_i} \lam_i^{y_i} e^{- h \sum_{d=1}^{y_i} t_{i,d}} \; 
 \prod_{i=1}^R \exp \left [ - \lam_i (1-e^{-h T}) \right] 
$$
where $\by$ is the vector of counts and $\bt$ the vector of times-to-each-detection, arranged conformably. The formula for constant $\lam$ follows immediately.

### R code


``` {r eval=FALSE}
# generates the times-to-each-detection for M4s with lambda constant
gM4sgen=function(param)
{
lam=param[1]
h=param[2]
yt=matrix(0,R,2);
for(i in 1:R)
{ n=rpois(1,lam)
if( n > 0)
{
  tvec=sort(rexp(n,h))
  tvecltT=as.matrix(tvec[tvec < Tmax])
  y=nrow(tvecltT)
  tsum=sum(tvecltT)
  yt[i,]=c(y,tsum)
}
}
return(yt)
}

# evaluates the log-likelihood for model M4s with lambda constant
gM4s=function(param)
{
lam=exp(param[1])
h=exp(param[2])
loglik=matrix(0,R,1)
tvec=gdat
for(i in 1:R)
{y=tvec[i,1]
tsum=tvec[i,2]
if(y == 0)
  {loglik[i]=loglik[i]-lam*(1-exp(-h*Tmax))}
else
  {loglik[i]=loglik[i]+y*log(h*lam)-lam*(1-exp(-h*Tmax))-h*tsum}
}
return(-sum(loglik))
}

# generates the times-to-each-detection for M4s with lambda site-dependent
xgM4sgen=function(param)
{
  b0=param[1]
  b1=param[2]
  h=param[3]
  lam=exp(b0+b1*vegHt)
  yt=matrix(0,R,2);
  for(i in 1:R)
  { n=rpois(1,lam[i])
  if( n > 0)
  {
    tvec=sort(rexp(n,h))
    tvecltT=as.matrix(tvec[tvec < Tmax])
    y=nrow(tvecltT)
    tsum=sum(tvecltT)
    yt[i,]=c(y,tsum)
  }
  }
  return(yt)
}

# evaluates the log-likelihood for model M4s with lambda site-dependent
xgM4s=function(param)
{
  b0=param[1]
  b1=param[2]
  h=exp(param[3])
  lam=exp(b0+b1*vegHt)
  loglik=matrix(0,R,1)
  tvec=gdatx
  for(i in 1:R)
  {
    y=tvec[i,1]
  tsum=tvec[i,2]
  if(y == 0)
  {loglik[i]=loglik[i]-lam[i]*(1-exp(-h*Tmax))}
  else
  {loglik[i]=loglik[i]+y*log(h*lam[i])-lam[i]*(1-exp(-h*Tmax))-h*tsum}
  }
  return(-sum(loglik))
}
```

## Model M4m: Times-to-each-detection with multiple visits 

### Formula 

For $R_t$ the number of sites at which times-to-each-detection are recorded, 
\begin{eqnarray*}
L_{M4m}( h,\lam_i;\by, \bt) & = &  \prod_{i=1}^{R_t} \prod_{j=1}^{J_{i,t}} \bigg[ \frac{h^{y_{ij}} \; e^{-h \sum_{d=1}^{y_{ij}} t_{i,jd}}}{\ds \left ( 1-e^{-h T} \right )^{y_{ij}}} \bigg] \times L_{M3m}(h,\lam_i; \by),
\end{eqnarray*}
where $\by$ is the vector of counts and $\bt$ the vector of times-to-each-detection, arranged conformably. This expression can be simplified by cancelling  the terms in $\left ( 1-e^{-h T} \right )$ to provide a computationally more efficient form, that is 
\begin{eqnarray*}
L_{M4m}(h,\lam_i;\by, \bt) & = &  \prod_{i=1}^{R_t} \bigg\{ \bigg( \prod_{j=1}^{J_{i,t}}  h^{y_{ij}} \; e^{-h \sum_{d=1}^{y_{ij}} t_{i,jd} } \bigg)\; e^{-\lam_{i}} \; \lam_i^{y_{i,M} } \; e^{- h T \sum_{j=1}^J (y_{i,M} -y_{ij})}  \\
& & \hspace{-1mm} 
   \times \hg{J-1}{J-1} 
\bigg( 
 \begin{array}{c}
(y_{i,M}+1)\\ 
(y_{i,M}-y_{ij}+1)
\end{array}
\bigg| \lam_i  e^{-h T J}) \bigg) \bigg\} \; \prod_{i=R_t+1}^R \exp[-\lam_i (1 - e^{-h T J})].
\end{eqnarray*}
The formulae for constant $\lam$ follow immediately.

## R code

``` {r eval=FALSE}

# generates times-to-each-detection for M4m with lambda constant
gM4mgen=function(param)
{
  lam=param[1]
  h=param[2]
  ymat=matrix(0,R,J)
  tsmat=matrix(0,R,J)
  for(i in 1:R)
  {n=rpois(1,lam)
  if(n==0){ymat[i,]=matrix(0,J,1)
  tsmat[i,]=matrix(0,J,1)}
  if( n > 0)
    for(j in 1:J)
    {
      tvec=sort(rexp(n,h))
      tvecltT=as.matrix(tvec[tvec < Tmax])
      ymat[i,j]=nrow(tvecltT)
      tsmat[i,j]=sum(tvecltT)
    }
  }
  return(cbind(ymat,tsmat))
}

# evaluates the log-likelihood for model M4m with lambda constant
gM4m=function(param)
{ ymat=gdat[,1:J]
  tsmat=gdat[,(J+1):(2*J)]
  lam=exp(param[1])
  h=exp(param[2])
loglik=0
for(i in 1:R)
{
  yvec=ymat[i,]
  yvec=sort(yvec, decreasing=FALSE)
  yij=yvec[-J]
  yiJ=yvec[J]
  tsum=tsmat[i,]
  term1=sum(yvec)*log(h)-h*sum(tsum)
  term2=-lam-h*sum(Tmax*(yiJ-yvec))+yiJ*log(lam)
  term3=(J-1)*log(factorial(yiJ))-sum(log(factorial(yiJ-yij)))
  theta=lam*exp(-h*Tmax*J)
  hgfun=genhypergeo(rep(yiJ,(J-1)) + 1,yiJ - yij + 1,theta)
  term4=log(hgfun)
  loglik=loglik+term1+term2+term3+term4}
return(-loglik)
}

# generates times-to-each-detection for M4m with lambda site-dependent
xgM4mgen=function(param)
{
  b0=param[1]
  b1=param[2]
  h=param[3]
  lam=exp(b0+b1*vegHt)
  ymat=matrix(0,R,J)
  tsmat=matrix(0,R,J)
  for(i in 1:R)
  { n=rpois(1,lam[i])
  if( n > 0)
    for(j in 1:J)
    {
      tvec=rexp(n,h)
      tvecltT=as.matrix(tvec[tvec < Tmax])
      ymat[i,j]=nrow(tvecltT)
      tsmat[i,j]=sum(tvecltT)
    }
  }
  return(cbind(ymat,tsmat))
}


# evaluates the log-likelihood for model M4m with lambda site-dependent
# y_j=0 then sum of arrival times is 0
xgM4m=function(param)
{ ymat=gdatx[,1:J]
  tsmat=gdatx[,(J+1):(2*J)]
  b0=param[1]
  b1=param[2]
  h=exp(param[3])
  lam=exp(b0+b1*vegHt)
  loglik=0
  for(i in 1:R)
  {
    yvec=ymat[i,]
    tsum=tsmat[i,]
    yvec=sort(yvec, decreasing=FALSE)
    yij=yvec[-J]
    yiJ=yvec[J]
    term1=sum(yvec)*log(h)-h*sum(tsum)
    term2=-lam[i]-h*sum(Tmax*(yiJ-yvec))+yiJ*log(lam[i])
    term3=(J-1)*log(factorial(yiJ))-sum(log(factorial(yiJ-yij)))
    theta=lam[i]*exp(-h*Tmax*J)
    hgfun=genhypergeo(rep(yiJ,(J-1)) + 1,yiJ - yij + 1,theta)
    term4=log(hgfun)
    loglik=loglik+term1+term2+term3+term4}
  return(-loglik)
}
```


## Model M5s: Time-to-first-detection and count data with a single visit 

### Formula 

For $R_t$ the number of sites at which counts and times-to-first detection are recorded, 
\begin{eqnarray*}
L_{M5s}(h,\lam_i; \by, \bt_{(1)}) & = & \prod_{i=1}^{R_t} \bigg[ h e^{-h t_{i(1)}} (e^{-h t_{i(1)}} -e^{- h T} )^{y_i-1} \bigg] \; \prod_{i=1}^R \exp \left [ - \lam_i (1-e^{-h T}) \right]
\end{eqnarray*}
where $\by$ and $\bt_{(1)}$ are vectors of counts and times, arranged conformably. The formula for constant $\lam$ follows immediately.


### R code

``` {r eval=FALSE}

# generates time-to-first-detection and count data for M5s with lambda constant
gM5sgen=function(param)
{
  lam=param[1]
  h=param[2]
  yt=matrix(0,R,2);
  for(i in 1:R)
  {n=rpois(1,lam)
  if( n > 0)
    {
    tvec=sort(rexp(n,h))
    tvecltT=as.matrix(tvec[tvec < Tmax])
    y=nrow(tvecltT)
    if(y >0)
     {t1=min(tvecltT)
      yt[i,]=c(y,t1)}
      }
  }
return(yt)
}

# evaluates the log-likelihood for model M5s with lambda constant
gM5s=function(param)
{
lam=exp(param[1])
h=exp(param[2])
ytmat=gdat
loglik=0
for(i in 1:R)
  {y=ytmat[i,1]
  t1=ytmat[i,2]
  if(y == 0)
    {loglik=loglik-lam*(1-exp(-h*Tmax))}else
    {term1=log(h)+y*log(lam)
     term2=(y-1)*log((exp(-h*t1)-exp(-h*Tmax)))
     term3=-lam*(1-exp(-h*Tmax))-h*t1
     term4=-log(factorial(y-1))
     loglik=loglik+term1+term2+term3+term4
}
}
return(-loglik)
}

# generates time-to-first-detection and count data for M5s with lambda site-dependent
xgM5sgen=function(param)
{
  b0=param[1]
  b1=param[2]
  h=param[3]
  lam=exp(b0+b1*vegHt)
  yt=matrix(0,R,2);
  for(i in 1:R)
  {n=rpois(1,lam[i])
  if( n > 0)
  {
    tvec=sort(rexp(n,h))
    tvecltT=as.matrix(tvec[tvec < Tmax])
    y=nrow(tvecltT)
    if(y >0)
      {t1=min(tvecltT)
       yt[i,]=c(y,t1)}
     }
  }
  return(yt)
}


# evaluates the log-likelihood for M5s with lambda site-dependent
xgM5s=function(param)
{ ytmat=gdatx
  b0=param[1]
  b1=param[2]
  h=exp(param[3])
  lam=exp(b0+b1*vegHt)
  loglik=0
  for(i in 1:R)
    {y=ytmat[i,1]
     t1=ytmat[i,2]
     if(y == 0)
  {loglik=loglik-lam[i]*(1-exp(-h*Tmax))}
     if(y >0)
  {term1=log(h)+y*log(lam[i])
   term2=(y-1)*log(exp(-h*t1)-exp(-h*Tmax))
   term3=-lam[i]*(1-exp(-h*Tmax))-h*t1
   term4=-log(factorial(y-1))
        loglik=loglik+term1+term2+term3+term4
       }
     }
  return(-loglik)
}
```
## Model M5m: Time-to-first-detection and count data with a single visit 

### Formula

For $R_t$ the number of sites and $J_{i,t}$ the number of visits to site $i$ at which  counts and times-to-first detection are recorded,
\begin{eqnarray*}
L_{M5m}(h,\lam_i;\by, \bt) & = &  \prod_{i=1}^{R_t} \bigg\{  \prod_{j=1}^{J_{i,t}}  \frac{h^{y_{ij}} \; e^{-h t_{ij(1)}} (e^{-h t_{ij(1)}} -e^{- h T} )^{y_{ij}-1}}{(1- e^{-h T})^{y_{ij}}} \bigg\} \times L_{M3m}(h,\lam_i; \by)
\end{eqnarray*}
where $\by$  and $\bt_{(1)}$ are vectors of counts and times, arranged conformably. This expression simplifies by cancelling terms in $(1-e^{-h T})$ as for model M4m and is not presented here. The formula for constant $\lam$ follows immediately.

### R code

``` {r eval=FALSE}

# generates time-to-first-detection and count data for M5m for lambda constant
gM5mgen=function(param)
{
  lam=param[1]
  h=param[2]
  ymat=matrix(0,R,J)
  t1mat=matrix(0,R,J)
  for(i in 1:R)
  { n=rpois(1,lam)
  if( n > 0)
    for(j in 1:J)
    {
      tvec=sort(rexp(n,h))
      tvecltT=as.matrix(tvec[tvec < Tmax])
      y=nrow(tvecltT)
      if(y==0)
      {ymat[i,j]=0
      t1mat[i,j]=0}else
      {ymat[i,j]=y
      t1mat[i,j]=min(tvecltT)}
    }
  }
  return(cbind(ymat,t1mat))
}

# evaluates the log-likelihood for M5m with lambda constant
gM5m=function(param)
{ ymat=gdat[,1:J]
  t1mat=gdat[,(J+1):(2*J)]
  lam=exp(param[1])
  h=exp(param[2])
  # all R
  loglik=0
  for(i in 1:R)
  {
    yvec=ymat[i,]
    if(sum(yvec)==0)
    {loglik=loglik-lam*(1-exp(-h*Tmax*J))}
    if(sum(yvec)>0)
    {# sort on yvec and t1
      t1vec=t1mat[i,]
      yt=cbind(yvec,t1vec)
      yt=yt[order(yt[,1],decreasing=FALSE),]
      yvec=yt[,1]
      t1vec=yt[,2]
      yij=yvec[-J]
      ymax=yvec[J]
      # terms for y[i,j] > 0 in site i
      for(j in 1:J)
        {y=yvec[j]
        if(y > 0)
        {term1=log(h)-h*t1vec[j]
        term2=(y-1)*log((exp(-h*t1vec[j])-exp(-h*Tmax)))
        term3=-sum(log(factorial(y-1)))
        loglik=loglik+term1+term2+term3}
        }
      #terms for all J equiv to sum over n
      term1=ymax*log(lam)
      term2=-lam-h*Tmax*sum(ymax-yvec)
      term3=(J-1)*log(factorial(ymax))-sum(log(factorial(ymax-yij)))
      theta=lam*exp(-h*Tmax*J)
      hgfun=genhypergeo(rep(ymax,(J-1)) + 1,ymax - yij + 1,theta)
      term4=log(hgfun)
      loglik=loglik+term1+term2+term3+term4}
  }
  return(-loglik)
}

# generates time-to-first-detection and count data for lambda site-dependent
xgM5mgen=function(param)
{
  b0=param[1]
  b1=param[2]
  h=param[3]
  lam=exp(b0+b1*vegHt)
  ymat=matrix(0,R,J)
  t1mat=matrix(0,R,J)
  for(i in 1:R)
  { n=rpois(1,lam[i])
  if( n > 0)
    for(j in 1:J)
    {
      tvec=sort(rexp(n,h))
      tvecltT=as.matrix(tvec[tvec < Tmax])
      y=nrow(tvecltT)
      if(y>0)
      {ymat[i,j]=y
      t1mat[i,j]=min(tvecltT)}
    }
  }
  return(cbind(ymat,t1mat))
}

# evaluates the log-likelihood for model M5m with lambda site-dependent
xgM5m=function(param)
{
  b0=param[1]
  b1=param[2]
  h=exp(param[3])
  ymat=gdatx[,1:J]
  t1mat=gdatx[,(J+1):(2*J)]
  lam=exp(b0+b1*vegHt)
  loglik=0
  for(i in 1:R)
  {
    yvec=ymat[i,]
    if(sum(yvec)==0)
    {loglik=loglik-lam[i]*(1-exp(-h*Tmax*J))}
    if(sum(yvec)>0)
    {# sort on yvec and t1
      t1vec=t1mat[i,]
      yt=cbind(yvec,t1vec)
      yt=yt[order(yt[,1],decreasing=FALSE),]
      yvec=yt[,1]
      t1vec=yt[,2]
      yij=yvec[-J]
      ymax=yvec[J]
      # terms for y[i,j] > 0 in site i
      for(j in 1:J)
        {y=yvec[j]
        if(y > 0)
        {term1=log(h)-h*t1vec[j]
        term2=(y-1)*log((exp(-h*t1vec[j])-exp(-h*Tmax)))
        term3=-sum(log(factorial(y-1)))
        loglik=loglik+term1+term2+term3}
        }
      #terms for all J equiv to sum over n
      term1=ymax*log(lam[i])
      term2=-lam[i]-h*Tmax*sum(ymax-yvec)
      term3=(J-1)*log(factorial(ymax))-sum(log(factorial(ymax-yij)))
      theta=lam[i]*exp(-h*Tmax*J)
      hgfun=genhypergeo(rep(ymax,(J-1)) + 1,ymax - yij + 1,theta)
      term4=log(hgfun)
      loglik=loglik+term1+term2+term3+term4}
  }
  return(-loglik)
}
```