---
output:
  pdf_document: default
  html_document: default
  word_document: default

---

# POISSON EXAMPLES: FUNCTIONS 
# Note: g is the Poisson rate parameter

# **PCount:S** 
# Pass yvec from generate to evaluate

# generate Poisson data for M3Ps with lambda constant
``` {r eval=FALSE}
gM3Psgen=function(param)
{lam=param[1]
g=param[2]
yvec=matrix(0,R,1)
for(i in 1:R)
{n=rpois(1,lam)
yvec[i]=rpois(1,g*n*Tmax)
}
return(yvec)
}
```

# evaluates the log-likelihood for M3Ps with lambda constant
``` {r eval=FALSE}
gM3Ps=function(param)
{lam=exp(param[1])
g=exp(param[2])
term=matrix(0,R,1)
mu=lam*exp(-g*Tmax)
for(i in 1:R)
{ y= yvec[i]
for(k in 0:y)
{
  term[i]=term[i]+exp(k*log(mu)+log(Stirling2(y,k)))}
}
loglik=-R*lam+R*lam*exp(-g*Tmax)+sum(yvec)*log(g*Tmax)+sum(log(term))
return(-loglik)
}
```

# generate Poisson data for M3Ps with covariate for lambda
``` {r eval=FALSE}
xgM3Psgen=function(param)
{b0=param[1]
b1=param[2]
g=param[3]
lam=exp(b0+b1*vegHt)
yvec=matrix(0,R,1)
for(i in 1:R)
{n=rpois(1,lam[i])
yvec[i]=rpois(1,g*n*Tmax)
}
return(yvec)
}
```

# evaluates the log-likelihood for M3Ps with lambda site dependent
``` {r eval=FALSE}
xgM3Ps=function(param)
{b0=param[1]
b1=param[2]
g=exp(param[3])
lam=exp(b0+b1*vegHt)
term=matrix(0,R,1)
mu=lam*exp(-g*Tmax)
for(i in 1:R)
{ y= yvec[i]
for(k in 0:y)
{
term[i]=term[i]+exp(k*log(mu[i])+log(Stirling2(y,k)))}
}
loglik=-sum(lam)+sum(mu)+sum(yvec)*log(g*Tmax)+sum(log(term))
return(-loglik)
}
```

# **PCount:M**
# Pass ymat from generate to evaluate

# generate Poisson data for M3Pm with lambda constant
``` {r eval=FALSE}
gM3Pmgen=function(param)
{lam=param[1]
g=param[2]
ymat=matrix(0,R,J)
for(i in 1:R)
{n=rpois(1,lam)
for(j in 1:J)
ymat[i,j]=rpois(1,g*n*Tmax)
}
return(ymat)
}
```

# calculate -log-ikelihood M3Pm with lambda constant
``` {r eval=FALSE}
gM3Pm=function(param)
{lam=exp(param[1])
g=exp(param[2])
mu=lam*exp(-g*Tmax*J)
fin=matrix(0,R,1)
term=matrix(0,R,1)
ytot=rowSums(ymat)
for(i in 1:R)
{ if(ytot[i] > 0)
{
  for(k in 0:ytot[i])
    {term[i]=term[i]+exp(k*log(mu)+log(Stirling2(ytot[i],k)))}
  fin[i]=fin[i]+log(term[i])
}
}
loglik=-R*lam+R*mu+sum(ymat)*log(g*Tmax)+sum(fin)
return(-loglik)
}
```

# generate Poisson data for M3Pm with lambda site dependent
``` {r eval=FALSE}
xgM3Pmgen=function(param)
{b0=param[1]
b1=param[2]
g=param[3]
lam=exp(b0+b1*vegHt)
ymat=matrix(0,R,J)
for(i in 1:R)
{n=rpois(1,lam[i])
for(j in 1:J)
  ymat[i,j]=rpois(1,g*n*Tmax)
}
return(ymat)
}
```

# calculate -log-ikelihood for M3Pm with lambda site-dependent
``` {r eval=FALSE}
xgM3Pm=function(param)
{b0=param[1]
b1=param[2]
g=exp(param[3])
lam=exp(b0+b1*vegHt)
fin=matrix(0,R,1)
term=matrix(0,R,1)
ytot=rowSums(ymat)
mu=lam*exp(-g*Tmax*J)
for(i in 1:R)
{if(ytot[i] > 0)
   {for(k in 0:ytot[i])
   {term[i]=term[i]+exp(k*log(mu[i])+log(Stirling2(ytot[i],k)))}
   fin[i]=fin[i]+log(term[i])}
}
loglik=-sum(lam)+sum(mu)+sum(ymat)*log(g*Tmax)+sum(fin)
return(-loglik)
}
```

# POISSON EXAMPLES: PROGRAMS

# For illustration: S with no covariates, M with covariates 

# PCount:S No covariates

## R changes to Rsites
## Tmax changes to Tsearch

``` {r eval=FALSE}
rm(list=ls())
library(copula)
# seed
set.seed(123)
# setting
R=100. #number of sites
Tmax=3 #maximum time
#true parameters
lamt=2
gt=0.4620981
# starting values and matrices for simulations 
nsim=1000
nprint=nsim/10
lamvec=array(0,c(nsim,1))
gvec=array(0,c(nsim,1))
paramt=c(lamt,gt)
param0=c(log(lamt),log(gt))
# cycle
for(isim in 1:nsim)
  { #generate data
  yvec=gM3Psgen(paramt)
  # optimize
  finopt=optim(param0,gM3Ps,method="BFGS")
  lamvec[isim,]=exp(finopt$par[1])
  gvec[isim,]=exp(finopt$par[2])
  #counter
  if(floor(isim/nprint)*nprint==isim)
  {print(isim)}
  }
```

# PCOUNT:M with covariates
## R changes to Rsites
## Tmax changes to Tsearch
## J changes to Jsites

``` {r eval=FALSE}
library(copula)
# seed
set.seed(123)
# setting
R=100. #number of sites
Tmax=3 #maximum time
J=5   #multisite site
#true parameters
b0t =0.660878  
b1t =0.255413
gt=0.4620981
# simulation of the covariate vegHt over R sites for true parameters
vegHt=rnorm(R,0,1)
hist(vegHt)
lambda=exp(b0t+b1t*vegHt)
mean(lambda)
# starting values and matrices for simulations 
nsim=1000
nprint=nsim/10
b0vec=array(0,c(nsim,1))
b1vec=array(0,c(nsim,1))
gvec=array(0,c(nsim,1))
paramt=c(b0t,b1t,gt)
param0=cbind(b0t,b1t,log(gt))
# cycle
for(isim in 1:nsim) {
  # generate
  ymat=xgM3Pmgen(paramt)
  # optimize
  finopt=optim(param0,xgM3Pm,method="BFGS")
  b0vec[isim,]=finopt$par[,1]
  b1vec[isim,]=finopt$par[,2]
  gvec[isim,]=exp(finopt$par[,3])
  #counter
  if(floor(isim/nprint)*nprint==isim)
  {print(isim)}
}
```