---
output:
  pdf_document: default
  html_document: default
  word_document: default
---
# Web Appendix for the paper ``Exact Likelihoods for N-mixture models with Time-to-Detection Data"

\def\blam{\mbox{\boldmath $\lambda$}}

# PROGRAMS

The R code provided in this section of the Web Appendix can be used to prepare programs for simulating binary, count and time-to-detection data based on the $N$-mixture models introduced in the main paper. The programs handle settings in which the abundance parameter is taken to be constant or site-dependent and the rate parameter and time search are held constant. Programs in which the rate parameter and the search time  are site- and visit-dependent are intricate and somewhat fragile for practically meaningful numbers of sites and are not included here.


## PROGRAMS FOR A CONSTANT ABUNDANCE PARAMETER

### SETTING

The R code below provides details of the study setting to be included at the start of all programs for which the abundance parameter is a constant, $\lambda$.

```{r eval=FALSE}
rm(list=ls())
library(copula)   #library for Stirling numbers of the second kind
library(hypergeo) #library for the generalised hypergeometric function
library(tictoc)
tic()
# setting
R=100  #number of sites
J=5    #number of visits: single J=1 OR multiple J > 1
Tmax=3 #maximum search time
#true parameters
lamt = 2.     #abundance
ht=0.4620981  #rate parameter
# true probabilites
1-exp(-ht*Tmax)      #P(individual detected)
1-exp(-lamt)         #P(site occupancy)
# starting values and matrices for the simulations
nsim=1000.                 #number of simulations
nprint=nsim/10             #progress printed 
lamvec=array(0,c(nsim,1))
hvec=array(0,c(nsim,1))
#parameter for the function generating the data
paramt=c(lamt,ht) 
#starting value for the working parameter in optim with lambda > 0 and h > 0
param0=cbind(log(lamt),log(ht)) 
```

### SIMULATION

The R code below provides details of the simulation of the model for time-to-first-detection data with multiple sites, that is model M2m. For other models, only the function names $\tt{gM2mgen}$ and $\tt{gM2m}$ need to be changed. The model functions are presented in the next section. The data generated are termed, generically, $\tt{gdat}$, for all models, irrespective of the dimensions of that data and are then accessed by the $\tt{optim}$ function.

```{r eval=FALSE}
#cycle
for(isim in 1:nsim)
{ 
  # generate data
  gdat=gM2mgen(paramt)
  # optimize
  finopt=optim(param0,gM2m,method="BFGS")
  lamvec[isim,]=exp(finopt$par[1])
  hvec[isim,]=exp(finopt$par[2])
  #counter
  if(floor(isim/nprint)*nprint==isim)
  {print(isim)}
}
```

### RESULTS

The R code below provides a summary of the $\tt{nsim}$ maximum likelihood estimates of the abundance parameter $\lambda$ and the rate parameter $h$ obtained by simulation.

```{r eval=FALSE}
# results for lambda
lamt
mean(lamvec)
signif(sd(lamvec),5)
summary(lamvec)
hist(lamvec,25)
boxplot(lamvec)

# results for h
ht
signif(mean(hvec),4)
signif(sd(hvec),4)
summary(hvec)
hist(hvec,25)
boxplot(hvec)

toc()

```

## PROGRAMS FOR SITE-DEPENDENT ABUNDANCE PARAMETERS

### SETTING

Here, for illustration purposes, the abundance parameters across the $R$ sites are generated from the distribution $\lambda=exp(\beta_{0}+\beta_{1} \mathrm{vegHt})$ where $\beta_{0}$ and $\beta_{1}$ are unknown parameters and vegHt is an independent standard normal variate. The R code otherwise follows that for constant abundance.

``` {r eval=FALSE}
rm(list=ls())
library(copula)
library(hypergeo)
library(tictoc)
tic()

# seed
set.seed(123)

# setting
R=100. #number of sites
J=5    #number of visits
Tmax=3 #maximum time

#true parameters
b0t =0.660878  #true intercept for log(lambda) 
b1t =0.255413. #true slope for log(lambda)
ht=0.4620981.  #rate parameter

# simulation of the covariate vegHt over R sites 
# hence the site-dependent abundance for the true parameters b0t and b1t
vegHt=rnorm(R,0,1)
hist(vegHt)
lambda=exp(b0t+b1t*vegHt)
mean(lambda)

# true probabilites
1-exp(-ht*Tmax)      #P(individual detected)
1-exp(-mean(lambda)) #P(site occupancy)

# starting values and matrices for the simulations
nsim=1000
nprint=nsim/10
b0vec=array(0,c(nsim,1))
b1vec=array(0,c(nsim,1))
hvec=array(0,c(nsim,1))
paramt=c(b0t,b1t,ht)
```

### SIMULATION


The R code below provides details of the simulation of the model for time-to-first-detection data with multiple sites and site-dependent abundance, that is model M2m. For other models, only the function names $\tt{xgM2mgen}$ and $\tt{xgM2m}$ need to be changed. The model functions are presented in the next section. The data so generated are termed, generically, $\tt{gdatx}$, for all such models to distinguish them from the data generated for constant abundance.

```{r eval=FALSE}
#cycle
set.seed(123)
for(isim in 1:nsim)
  { #generate data
  gmatm=xgM2mgen(paramt)
   # optimize
  param0=c(b0t,b1t,log(ht))
  finopt=optim(param0,xgM2m,method="BFGS")
  b0vec[isim] =finopt$par[1] 
  b1vec[isim] =finopt$par[2] 
  hvec[isim]=exp(finopt$par[3])
  #counter
  if(floor(isim/nprint)*nprint==isim)
  {print(isim)}
}
```

### RESULTS

The R code below provides a summary of the $\tt{nsim}$ maximum likelihood estimates of the unknown parameters defining site abundance, $\beta_0$ and $\beta_1$, and the rate parameter $h$ obtained by simulation.

```{r eval=FALSE}
# results for b0
b0t
signif(mean(b0vec),4)
signif(sd(b0vec),4)
summary(b0vec)
hist(b0vec,25)
boxplot(b0vec)

# results for b1
b1t
signif(mean(b1vec),4)
signif(sd(b1vec),4)
summary(b1vec)
hist(b1vec,25)
boxplot(b1vec)

# results for ht
ht
signif(mean(hvec),4)
signif(sd(hvec),4)
summary(hvec)
hist(hvec,25)
boxplot(hvec)

toc()
```
